---
title: "chapter 1"
author: "Harrison Linder"
date: "`r Sys.Date()`"
output: html_document
---

below are the necessary libraries to use for this course

```{r}
library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)

options(scipen=999)
options(tigris_class = "sf")

root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")
```

Here is the color palette we are going to use for maps
```{r}
palette5 <- c("#f0f9e8","#bae4bc","#7bccc4","#43a2ca","#0868ac")

```

how to install my census api key
```{r}
census_api_key("c08e727b3d32b823955aea424744c9dda3b30df0", install = TRUE, overwrite = TRUE)
# readRenviron("~/.Renviron") is for some reason needed for the api to be active in the current session 
readRenviron("~/.Renviron")
```
we will be using the following data

from the textbook: In this chapter, the tidycensus is used to query and return Census data directly into R. At the time of publication however, the Census has disabled the year 2000 endpoint/data.10 Instead, the code below downloads an equivalent data frame called tracts00. A data frame is the most common way to store data in R. The code block at the end of this section demonstrates a tidycensus call to return 2017 data.
```{r}
tracts00 <- 
  st_read(file.path("PHL_CT00.geojson")) %>% 
    st_transform('ESRI:102728')
```

